---
title: "Creation of grp files for visualization"
author:
- affiliation: AG Annegret Wilde, Institute of Biology III
  name: "Ute Hoffmann"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    fig_width: 15
    fig_height: 8
    theme: united
    toc: yes
    number_sections: true
  pdf_document:
    toc: yes
    number_sections: true
---

```{r setup, include=FALSE, echo=FALSE}
require("knitr")
opts_knit$set(root.dir = "/data/Dokumente/uni/Doktorarbeit/T161V_analyses/00_Manuscript/Code/4_Multireads/")
library(tidyverse)
library(DESeq2)
library(edgeR)
library(ggplot2)
library(ggrepel)
library(ggpubr)
library(pheatmap)
library(RColorBrewer)
library(vsn)
library(GenomicRanges)
library(Biostrings)
library(rtracklayer)
library(viridis)
```

```{r}
#import gff for anno-TSS
features <- rtracklayer::import("Input/20210217_syne_onlyUnique_withFeat.gff3")
TUs <- rtracklayer::import("Input/Kopf_4091_TUs_combined.gff3")
```

```{r, echo=FALSE}
# some functions which are incredibly handy

#### Functions for plotting and handling DESeq2 objects

count_up_down <- function(deseq2_dframe, foldchange=1, padjusted=0.01)
{ # input: DESeq2 results object / data frame; foldchange and padjusted; returns numbers of features in DESeq2 results object differentially expressed according to cut-offs
  deseq2_dframe <- as.data.frame(subset(deseq2_dframe, !is.na(deseq2_dframe$padj)))
  # create vectors of down- or upregulated features, give out their numbers
  down <- deseq2_dframe$log2FoldChange < -foldchange & deseq2_dframe$padj < padjusted
  up <- deseq2_dframe$log2FoldChange > foldchange & deseq2_dframe$padj < padjusted
  
  print(paste("number of features down: ", sum(down)))
  print(paste("number of features up: ", sum(up)))
}

return_up_down <- function(deseq2_dframe, foldchange=1, padjusted=0.01)
{
  deseq2_dframe <- subset(deseq2_dframe, !is.na(deseq2_dframe$padj))
  # create vectors of down- or upregulated features, give out their numbers
  down <- deseq2_dframe$log2FoldChange < -foldchange & deseq2_dframe$padj < padjusted
  up <- deseq2_dframe$log2FoldChange > foldchange & deseq2_dframe$padj < padjusted
  
  vector_differentiallyExpressed <- c(row.names(deseq2_dframe)[down],row.names(deseq2_dframe)[up])
  return(vector_differentiallyExpressed)
}

volcanoPlot_ggplot <-  function(deseq2_dframe, foldchange=1, padjusted=0.01, color=TRUE, text=FALSE, numbers=TRUE, lines_padj_FC=TRUE)
{ # plots Volcano Plot for DESeq2 results object
  # compare: https://biocorecrg.github.io/CRG_RIntroduction/volcano-plots.html
  
  # The significantly differentially expressed genes are the ones found in the upper-left and upper-right corners.
  # Add a column to the data frame to specify if they are UP- or DOWN- regulated (log2FoldChange respectively positive or negative)
  
  # add a column of NAs
  df_copy <- as.data.frame(deseq2_dframe)
  df_copy$diffexpressed <- "NO"
  # if log2Foldchange > 0.6 and pvalue < 0.05, set as "UP" 
  df_copy$diffexpressed[df_copy$log2FoldChange > foldchange & df_copy$padj < padjusted] <- "UP"
  # if log2Foldchange < -0.6 and pvalue < 0.05, set as "DOWN"
  df_copy$diffexpressed[df_copy$log2FoldChange < -foldchange & df_copy$padj < padjusted] <- "DOWN"
  
  # prepare labels of plot
  df_copy$delabel <- NA
  df_copy$delabel[df_copy$diffexpressed != "NO"] <- row.names(df_copy)[df_copy$diffexpressed != "NO"]
  
  # Plot
  p <- ggplot(data=df_copy, aes(x=log2FoldChange, y=-log10(padj), col=diffexpressed, label=delabel)) + geom_point(alpha=0.3, show.legend = FALSE) + 
    theme_light() + labs(y="-Log10(p.adj_BH)", x="Log2FC") + theme(legend.position = "none")
  
  # color points
  mycolors <- c("#e69f00b2","#005a96b2",  "#d3d3d3b2")
  names(mycolors) <- c("DOWN", "UP", "NO")
  p <- p + scale_colour_manual(values = mycolors)
  
  # Add vertical lines for log2FoldChange thresholds, and one horizontal line for the p-value threshold 
  if(lines_padj_FC){
    p <- p + geom_vline(xintercept=c(-foldchange, foldchange), linetype="dotted") +
      geom_hline(yintercept=-log10(padjusted), linetype="dotted") # http://www.sthda.com/english/wiki/ggplot2-line-types-how-to-change-line-types-of-a-graph-in-r-software
  }
  
  # Now write down the name of genes beside the points...
  # Create a new column "delabel" to de, that will contain the name of genes differentially expressed (NA in case they are not)
  if(text){
    p <- p + geom_text_repel(fontface="italic")   #geom_text()
  }
  
  return(p)
}

MAplot_ggplot <-  function(deseq2_dframe, foldchange=1, padjusted=0.05, color=TRUE, text=FALSE, numbers=TRUE, lines_padj_FC=TRUE, y_axis_label="Log2FC")
{ # plots MAplot for DESeq2 results object
  # add a column of NAs
  df_copy <- as.data.frame(deseq2_dframe)
  df_copy$diffexpressed <- "NO"
  # if log2Foldchange > 0 and pvalue < 0.05, set as "UP" 
  df_copy$diffexpressed[df_copy$log2FoldChange > 0 & df_copy$padj < padjusted] <- "UP"
  # if log2Foldchange < 0 and pvalue < 0.05, set as "DOWN"
  df_copy$diffexpressed[df_copy$log2FoldChange < 0 & df_copy$padj < padjusted] <- "DOWN"
  
  # prepare labels of plot
  df_copy$delabel <- NA
  df_copy$delabel[df_copy$diffexpressed != "NO"] <- row.names(df_copy)[df_copy$diffexpressed != "NO"]
  
  # Plot
  p <- ggplot(data=df_copy, aes(x=baseMean, y=log2FoldChange, col=diffexpressed, label=delabel)) + geom_point(size=0.5, show.legend = FALSE) + 
    theme_light() + labs(y=y_axis_label, x="Mean of Normalized Counts") + theme(legend.position = "none") + scale_x_continuous(trans='log10')
  
  # color points
  mycolors <- c("#e69f00ff","#005a96ff",  "#d3d3d3ff")
  names(mycolors) <- c("DOWN", "UP", "NO")
  p <- p + scale_colour_manual(values = mycolors)
  
  # Add vertical lines for log2FoldChange thresholds, and one horizontal line for the p-value threshold 
  if(lines_padj_FC){
    p <- p + geom_hline(yintercept=c(-foldchange, 0, foldchange), linetype="dotted")
  }
  
  # Now write down the name of genes beside the points...
  # Create a new column "delabel" to de, that will contain the name of genes differentially expressed (NA in case they are not)
  if(text){
    p <- p + geom_text_repel(fontface="italic")   #geom_text()
  }
  
  return(p)
}

PCA_plot <- function(dds, title){
  # plots PCA for DESeq2 Data Set object, uses rlog, assumes DESeq2 data set object for dWT / WT / TV comparison
  rld <- rlog(dds)
  pcaData <- plotPCA(rld, intgroup=c("strain"), returnData=TRUE)
  percentVar <- round(100 * attr(pcaData, "percentVar"))
  ggplot(pcaData, aes(PC1, PC2, color=strain)) +
    geom_point(size=3) +
    xlab(paste0("PC1: ",percentVar[1],"% variance")) +
    ylab(paste0("PC2: ",percentVar[2],"% variance")) + 
    coord_fixed() + theme_light() + scale_color_manual(name="Strain", values=c("#000000ff", "#005a96ff", "#e69f00ff"), breaks=c("WT", "dWT", "TV"), labels=c("WT", "rne(WT)", "rne(5p)")) + 
    theme(legend.position="bottom", legend.box="vertical", legend.margin=margin(t=-0.2, r=0, b=0, l=0, unit="cm"))  + 
    labs(title=title)
}

heatmap_plot <- function(dds, title){
  # plots heat map for DESeq2 Data Set object, uses rlog
  dat <- assay(rlog(dds))
  dists_rl <- dist(t(dat))
  mat <- as.matrix(dists_rl)
  #rownames(mat) <- c(rep("rne(WT)", 3),rep("WT", 3), rep("TV",2))
  colnames(mat) <- NULL
  colors <- colorRampPalette(rev(brewer.pal(9, "Blues")))(255)
  pheatmap(
    mat,
    clustering_distance_rows = dists_rl,
    clustering_distance_cols = dists_rl,
    col = colors,
    main = "Sample-to-sample distances"
  )
}

PCA_plot_PSSTSS <- function(dds, title){
  # plots PCA for PSS / TSS comparison of DESeq Data set
  rld <- rlog(dds)
  pcaData <- plotPCA(rld, intgroup=c("type", "strain"), returnData=TRUE)
  percentVar <- round(100 * attr(pcaData, "percentVar"))
  return(ggplot(pcaData, aes(PC1, PC2, color=strain, shape=type)) +
    geom_point(size=3) +
    xlab(paste0("PC1: ",percentVar[1],"% variance")) +
    ylab(paste0("PC2: ",percentVar[2],"% variance")) + 
    coord_fixed() + theme_light() + scale_color_manual(name="Strain", values=c("#000000ff", "#005a96ff", "#e69f00ff"), breaks=c("WT", "dWT", "TV"), labels=c("WT", "rne(WT)", "rne(5p)")) +
    theme(legend.position="bottom", legend.box="vertical", legend.margin=margin(t=-0.2, r=0, b=0, l=0, unit="cm")) +
    labs(title=title) + scale_shape_discrete(name="", breaks=c("PSS", "TSS"), labels=c("PSS", "TSS")))
}

heatmap_plot_PSSTSS <- function(dds, title){
  # plots heat map for PSS / TSS comparison of DESeq Data set
  dat <- assay(rlog(dds))
  dists_rl <- dist(t(dat))
  mat <- as.matrix(dists_rl)
  #rownames(mat) <- c(rep(c("rne(5p) 0h", "rne(5p) 1h"), 3),rep(c("rne(WT) 0h", "rne(WT) 1h"), 3))
  colnames(mat) <- NULL
  colors <- colorRampPalette(rev(brewer.pal(9, "Blues")))(255)
  pheatmap(
    mat,
    clustering_distance_rows = dists_rl,
    clustering_distance_cols = dists_rl,
    col = colors,
    main = "Sample-to-sample distances"
  )
}

pvaluePlot <- function(result_object, title_plot){ 
  # plots histogram of distribution of p values in DESeq2 results object
  use <- result_object$baseMean > metadata(result_object)$filterThreshold
  h1 <- hist(result_object$pvalue[!use], breaks=0:50/50, plot=FALSE)
  h2 <- hist(result_object$pvalue[use], breaks=0:50/50, plot=FALSE)
  colori <- c(`filtered (low count)`="khaki", `not filtered`="powderblue")
  barplot(height = rbind(h1$counts, h2$counts), beside = FALSE,
        col = colori, space = 0, main = paste("Histogram of p-values for", title_plot), ylab="Frequency", xlab="P-values")
  text(x = c(0, length(h1$counts)), y = 0, label = paste(c(0,1)),
     adj = c(0.5,1.7), xpd=NA)
  legend("topright", fill=rev(colori), legend=rev(names(colori)))
}

#### Functions to work on GRanges objects
create_GRanges_object <- function(pos_list, base_means){
  # creates GRanges object from pos_list and adds additional info about base_means (base_means input as vector)
  # input: pos_list: List of positions which should be used for GRanges object: each element in list: [1]: contig name, [2]: nt-position (=start/stop), [3]: strand information
  # with some extra constructs (e.g. as.integer(map_chr(pos_list, pluck, 2)) one can work around for-construct, but it's slower https://rstudio-education.github.io/tidyverse-cookbook/transform-lists-and-vectors.html
  
  # initialise vectors
  seq_names <- c()
  starts <- c()
  stops <- c()
  strands <- c()

  for(i in 1:length(pos_list)){
    seq_names[i] <- pos_list[[i]][1]
    starts[i] <- as.integer(pos_list[[i]][2])
    stops[i] <- as.integer(pos_list[[i]][2])
    if(pos_list[[i]][3]=="plus"){
      strands[i] <- "+"
    }else if(pos_list[[i]][3]=="minus"){
      strands[i] <- "-"
    }
  }
  
  GRanges_object <- GRanges(seqnames=seq_names,
      ranges=IRanges(starts, stops), strand=strands, baseMean=base_means)

  return(GRanges_object)
}

create_GRanges_object_from_resObject <- function(DESeq2_resObject, foldchange, padjusted, up=TRUE){
  # input: DESeq2 results object, + foldchange and padjusted cut-offs which should be used to only select certain set of positions, up=TRUE -> extract positions with log2FC > foldchange, else (up=FALSE) -> extract positions with log2FC < foldchange 
  if(up){
  subset_resObject <- subset(DESeq2_resObject, DESeq2_resObject$log2FoldChange>foldchange & DESeq2_resObject$padj < padjusted)
  } else{
  subset_resObject <- subset(DESeq2_resObject, DESeq2_resObject$log2FoldChange<foldchange & DESeq2_resObject$padj < padjusted)
  }
  positions<- base::strsplit(row.names(subset_resObject), "-")
  rangesObject <- create_GRanges_object(positions, subset_resObject$baseMean)
  return(rangesObject)
}

save_gff <- function(Ranges, file_path, basic_name){
  rtracklayer::export(Ranges, paste(file_path, basic_name, "_all.gff", sep=""))
  rtracklayer::export(subset(Ranges, seqnames(Ranges)=="BA000022.2"), paste(file_path, basic_name, "_chromo.gff", sep=""))
  rtracklayer::export(subset(Ranges, seqnames(Ranges)=="AP004311.1"), paste(file_path, basic_name, "_pSYSA.gff", sep=""))
  rtracklayer::export(subset(Ranges, seqnames(Ranges)=="AP004310.1"), paste(file_path, basic_name, "_pSYSM.gff", sep=""))
  rtracklayer::export(subset(Ranges, seqnames(Ranges)=="AP004312.1"), paste(file_path, basic_name, "_pSYSG.gff", sep=""))
  rtracklayer::export(subset(Ranges, seqnames(Ranges)=="AP006585.1"), paste(file_path, basic_name, "_pSYSX.gff", sep=""))
}

advanced_reduce <- function(rangesObject){
  # function to reduce rangesObject (join adjacent positions to one longer position), and at same time trim positions with width > 0 to the one with highest population of reads (highest baseMean)
  new_ranges <- GenomicRanges::reduce(rangesObject)
  redundant_pos <- new_ranges[which(width(new_ranges)>1)]
  
  indices_new_ranges <- which(width(new_ranges)>1)
  
  starts <- start(new_ranges)
  ends <- end(new_ranges)
  strands <- strand(new_ranges)
  seqname_vec <- seqnames(new_ranges)
  
  for(i in indices_new_ranges){
      # extract baseMeans from original object    
      ranges_index <- which(start(rangesObject) %in% starts[i]:ends[i] & strand(rangesObject)==strands[i] & seqnames(rangesObject)==seqname_vec[i])
      baseMeans <- rangesObject[ranges_index]$baseMean
      positions <- starts[i]:ends[i]
  
      start(new_ranges[i]) <- positions[which(baseMeans==max(baseMeans))]
      end(new_ranges[i]) <- positions[which(baseMeans==max(baseMeans))]
  }

  return(new_ranges)
}

advanced_reduce_withBaseMeans <- function(rangesObject){
  # function to reduce rangesObject (join adjacent positions to one longer position), and at same time trim positions with width > 0 to the one with highest population of reads (highest baseMean)
  new_ranges <- GenomicRanges::reduce(rangesObject)
  new_ranges$baseMeans <- NA
  redundant_pos <- new_ranges[which(width(new_ranges)>1)]
  
  indices_new_ranges <- which(width(new_ranges)>1)
  indices_new_ranges_width0 <- which(width(new_ranges)==1)
  
  starts <- start(new_ranges)
  ends <- end(new_ranges)
  strands <- strand(new_ranges)
  seqname_vec <- seqnames(new_ranges)
  
  for(i in 1:length(new_ranges)){
      # extract baseMeans from original object    
      ranges_index <- which(start(rangesObject) %in% starts[i]:ends[i] & strand(rangesObject)==strands[i] & seqnames(rangesObject)==seqname_vec[i])
      baseMeans <- rangesObject[ranges_index]$baseMean
      positions <- starts[i]:ends[i]
  
      start(new_ranges[i]) <- positions[which(baseMeans==max(baseMeans))]
      end(new_ranges[i]) <- positions[which(baseMeans==max(baseMeans))]
      new_ranges[i]$baseMeans <- max(baseMeans)
  }

  return(new_ranges)
}
```

# Create .grps

Coverage data was created using Filtered .bam files with multireads, which were processed by bedtools genome coverage. Files were re-named in the respective history and downloaded as one .zip. In a next step, separate files were combined into one large table using Python code (joinFiles_PSS.py, joinFiles_TSS.py, joinFiles_transcript.py).

Normalization factors are size factors determined DESeq2 using separate .Rmd files (PSS-TSS Analysis.Rmd, TranscriptAnalysis.Rmd).

```{r}
transcript_factors <- read.table("Input/transcript_sizeFactors.csv", sep=",", header=TRUE, row.names=1)
TSS_factors <- read.table("Input/TSS_sizeFactors.csv", sep=",", header=TRUE, row.names=1)
PSS_factors <- read.table("Input/PSS_sizeFactors.csv", sep=",", header=TRUE, row.names=1)
row.names(TSS_factors) <- c("dWT1", "dWT2", "dWT3", "WT1", "WT2", "WT3", "TV1", "TV2")
row.names(PSS_factors) <- c("dWT1", "dWT2", "dWT3", "WT1", "WT2", "WT3", "TV1", "TV2")
coldata_general <- read.table("Input/colData.csv", sep=",", header=TRUE, row.names=1)
```

## PSS data

```{r}
PSS_coverage <- read.table("Input/multireads_PSS_5ends_combined_5sensing.txt", sep="\t", header=TRUE, row.names=1)
PSS_fact <- PSS_factors$factor

if(FALSE){
ddsMat_PSS <- DESeqDataSetFromMatrix(countData = PSS_coverage,
                                 colData = coldata_general,
                                 design = ~ strain)

names(PSS_fact) <- row.names(PSS_factors)
ddsMat_PSS$sizeFactor <- PSS_fact
  
PSS_norm_counts <- counts(ddsMat_PSS, normalized=TRUE)
grp_File <- data.frame("WT"=apply(PSS_norm_counts[,c("WT1", "WT2", "WT3")],1,mean), "dWT"=apply(PSS_norm_counts[,c("dWT1", "dWT2", "dWT3")],1,mean), "TV"=apply(PSS_norm_counts[,c("TV1", "TV2")],1,mean))
 
grp_File$seqname <- rep(NA, length(grp_File$WT))
seqnames <- c("BA000022.2", "AP004310.1", "AP004311.1", "AP004312.1", "AP006585.1")
for(i in seqnames){
  grp_File[which(grepl(i, row.names(grp_File))),]$seqname <- i
}
grp_File$strand <- rep(NA, length(grp_File$WT))
for(i in c("plus", "minus")){
  grp_File[which(grepl(i, row.names(grp_File))),]$strand <- i
}

for(i in seqnames){
  for(j in c("plus", "minus")){
    tmp <- subset(grp_File, grp_File$seqname==i & grp_File$strand==j)[,c("WT", "dWT", "TV")]
    s=""
    if(j=="plus"){
      s="fw"
    }else{
      s="rev"
   }
    write.table(tmp, file=paste("Output/grp/PSS/", i, "_PSS-multireads_WT-dWT-TV_", s, ".grp", sep=""), sep="\t", row.names=FALSE, col.names = FALSE)
  }
}
  
rm(PSS_norm_counts)
}

```

## TSS data

```{r}
TSS_coverage <- read.table("Input/multireads_TSS_5ends_combined_5sensing.txt", sep="\t", header=TRUE, row.names=1)
ddsMat_TSS <- DESeqDataSetFromMatrix(countData = TSS_coverage,
                                 colData = coldata_general,
                                 design = ~ strain)
TSS_fact <- TSS_factors$factor
names(TSS_fact) <- row.names(TSS_factors)
ddsMat_TSS$sizeFactor <- TSS_fact

TSS_norm_counts <- counts(ddsMat_TSS, normalized=TRUE)
rm(ddsMat_TSS)

if(FALSE){
grp_File <- data.frame("WT"=apply(TSS_norm_counts[,c("WT1", "WT2", "WT3")],1,mean), "dWT"=apply(TSS_norm_counts[,c("dWT1", "dWT2", "dWT3")],1,mean), "TV"=apply(TSS_norm_counts[,c("TV1", "TV2")],1,mean))
 
grp_File$seqname <- rep(NA, length(grp_File$WT))
seqnames <- c("BA000022.2", "AP004310.1", "AP004311.1", "AP004312.1", "AP006585.1")
for(i in seqnames){
  grp_File[which(grepl(i, row.names(grp_File))),]$seqname <- i
}
grp_File$strand <- rep(NA, length(grp_File$WT))
for(i in c("plus", "minus")){
  grp_File[which(grepl(i, row.names(grp_File))),]$strand <- i
}

for(i in seqnames){
  for(j in c("plus", "minus")){
    tmp <- subset(grp_File, grp_File$seqname==i & grp_File$strand==j)[,c("WT", "dWT", "TV")]
    s=""
    if(j=="plus"){
      s="fw"
    }else{
      s="rev"
   }
    write.table(tmp, file=paste("Output/grp/TSS/", i, "_TSS-multireads_WT-dWT-TV_", s, ".grp", sep=""), sep="\t", row.names=FALSE, col.names = FALSE)
  }
}
  
}
```

## Transcript data

```{r}
transcript_coverage <- read.table("Input/multireads_transcript_coverage_combined_5sensing.txt", sep="\t", header=TRUE, row.names=1)
ddsMat_transcript <- DESeqDataSetFromMatrix(countData = transcript_coverage,
                                 colData = coldata_general,
                                 design = ~ strain)
transcript_fact <- transcript_factors$factor
names(transcript_fact) <- row.names(transcript_factors)
ddsMat_transcript$sizeFactor <- transcript_fact

transcript_norm_counts <- counts(ddsMat_transcript, normalized=TRUE)
rm(ddsMat_transcript)
rm(transcript_coverage)

if(FALSE){
  grp_File <- data.frame("WT"=apply(transcript_norm_counts[,c("WT1", "WT2", "WT3")],1,mean), "dWT"=apply(transcript_norm_counts[,c("dWT1", "dWT2", "dWT3")],1,mean), "TV"=apply(transcript_norm_counts[,c("TV1", "TV2")],1,mean))
  grp_File$seqname <- rep(NA, length(grp_File$WT))
  seqnames <- c("BA000022.2", "AP004310.1", "AP004311.1", "AP004312.1", "AP006585.1")
  for(i in seqnames){
    grp_File[which(grepl(i, row.names(grp_File))),]$seqname <- i
  }
  grp_File$strand <- rep(NA, length(grp_File$WT))
  for(i in c("plus", "minus")){
    grp_File[which(grepl(i, row.names(grp_File))),]$strand <- i
  }
  for(i in seqnames){
    for(j in c("plus", "minus")){
      tmp <- subset(grp_File, grp_File$seqname==i & grp_File$strand==j)[,c("WT", "dWT", "TV")]
      s=""
      if(j=="plus"){
        s="fw"
      }else{
        s="rev"
      }
      write.table(tmp, file=paste("Output/grp/transcript/", i, "_transcript-multireads_WT-dWT-TV_", s, ".grp", sep=""), sep="\t", row.names=FALSE, col.names = FALSE)
    }
  }
}
```

# Perform DESeq2 analysis

```{r read-Input}
names(PSS_coverage) <- paste(names(PSS_coverage), "_PSS", sep="")
names(PSS_fact) <- paste(names(PSS_fact), "_PSS", sep="")
names(TSS_coverage) <- paste(names(TSS_coverage), "_TSS", sep="")
names(TSS_fact) <- paste(names(TSS_fact), "_TSS", sep="")

# merge PSS and TSS to merged_raw
merged_raw <- merge(PSS_coverage, TSS_coverage, by="row.names")
row.names(merged_raw) <- merged_raw$Row.names
merged_raw$Row.names <- NULL
rm(TSS_coverage)
```

When taking into account lowly populated sites, DESeq2 dispersion estimates become over-fitted. Hence, edgeR::filterByExpression() is used as a pre-processing step before further DESeq2 analyses.

```{r edgeR-filter}
group=c(rep("dWT_PSS", 3), rep("WT_PSS", 3), rep("TV_PSS", 2), rep("dWT_TSS", 3), rep("WT_TSS", 3), rep("TV_TSS", 2))
y <- DGEList(counts=merged_raw, group=group)
nrow(y)
keep <- filterByExpr(y)
y <- y[keep, ,keep.lib.size=FALSE]
nrow(y)

merged_filtered <- merged_raw[row.names(y$counts),]
nrow(merged_filtered)
rm(merged_raw)

coldata <- read.csv("Input/colData_PSS-TSS.csv", row.names=1)

# create DESeq2 data object
ddsMat_PSSTSS <- DESeqDataSetFromMatrix(countData = merged_filtered,
                                 colData = coldata,
                                 design = ~ strain + type)

ddsMat_PSSTSS$sizeFactor <- c(PSS_fact, TSS_fact)
ddsMat_PSSTSS <- DESeq(ddsMat_PSSTSS)

res_PSSTSS <- results(ddsMat_PSSTSS, contrast=c("type", "PSS", "TSS"))
write.csv(res_PSSTSS[order(res_PSSTSS$padj),], file="Output/DESeq2_resultsTables/results_PSS-TSS-comparisons.csv")
```

## Diagnostic Plots

```{r}
plotDispEsts(ddsMat_PSSTSS)
pdf(file="Output/DESeq2_Plots/ddsMat_PSSTSS_DispEsts.pdf", width=4.5, height=4.5)
plotDispEsts(ddsMat_PSSTSS, xlab="Mean of Normalized Counts", ylab="Dispersion")
dev.off()

p <- PCA_plot_PSSTSS(ddsMat_PSSTSS, "PSS-TSS")
p
ggsave("Output/DESeq2_Plots/ddsMat_PSSTSS_PCA.pdf", plot=p, width=15, height=9, units="cm")

p <- heatmap_plot_PSSTSS(ddsMat_PSSTSS, "PSS-TSS")
p
ggsave("Output/DESeq2_Plots/ddsMat_PSSTSS_heatMap.pdf", plot=p, width=15, height=12, units="cm")
rm(ddsMat_PSSTSS) # to free space

pdf(file="Output/DESeq2_Plots/ddsMat_PSSTSS_hist_log2FC.pdf", width=4.5, height=4.5)
hist(res_PSSTSS$log2FoldChange, breaks=20, main="", xlab="Log2FC(PSS/TSS)", ylab="Frequency")
dev.off()
hist(res_PSSTSS$log2FoldChange, breaks=20, main="DESeq2 Normalization", xlab="Log2FC PSS/TSS", ylab="Frequency")

# non-normalized
log2_all <- log2(apply(merged_filtered[,1:8],1, mean)/apply(merged_filtered[,9:16],1,mean))
hist(log2_all, breaks=20, main="No normalization", xlab="Log2(PSS/TSS)")
rm(merged_filtered)

p <- MAplot_ggplot(res_PSSTSS, foldchange=0.8, y_axis_label = "Log2 fold-change(PSS/TSS)")
p <- p + scale_colour_manual(values=c("UP"="#4e9879ff", "DOWN"="#9b511fff", "NO"="#d3d3d3ff"))
p
ggsave("Output/DESeq2_Plots/ddsMat_PSSTSS_MAplot.pdf",plot=p, width=15, height=12, units="cm")
```

## Extract DESeq2 results

```{r PSSTSS-extract-DESeq2-results}
res_PSSTSS <- subset(res_PSSTSS, !is.na(res_PSSTSS$padj))

# down: TSS, up: PSS
count_up_down(res_PSSTSS, foldchange=0.8, padjusted=0.05)
p <- volcanoPlot_ggplot(res_PSSTSS, foldchange=0.8, padjusted=0.05)
p
ggsave("Output/DESeq2_Plots/ddsMat_PSSTSS_VolcanoPlot.pdf", plot=p, width=15, height=12, units="cm")

pdf(file="Output/DESeq2_Plots/ddsMat_PSSTSS_pValuePlot.pdf", width=4.5, height=4.5)
pvaluePlot(res_PSSTSS, "Comparison PSS to TSS")
dev.off()
pvaluePlot(res_PSSTSS, "Comparison PSS to TSS")

# extract which positions are TSS, which are PSS
PSS_positions_unfiltered <- row.names(subset(res_PSSTSS, res_PSSTSS$log2FoldChange>0.8 & res_PSSTSS$padj<0.05))
PSS_nonReduced_Ranges <- create_GRanges_object(base::strsplit(PSS_positions_unfiltered, "-"), res_PSSTSS[PSS_positions_unfiltered,]$baseMean)
TSS_positions_unfiltered <- row.names(subset(res_PSSTSS, res_PSSTSS$log2FoldChange<(-0.8) & res_PSSTSS$padj<0.05))
```

## Filter for ratio 5' ends / transcript coverage

```{r filterTSS-positions}
trans_norm <- as.data.frame(transcript_norm_counts)
rm(transcript_norm_counts)
TSS_norm <- as.data.frame(TSS_norm_counts)
rm(TSS_norm_counts)
```

```{r}
TSS_trans_ratio_cutoff <- 0.02
TSS_trans_ratio <- apply(TSS_norm[TSS_positions_unfiltered,],1,mean)/apply(trans_norm[TSS_positions_unfiltered,],1,mean)
TSS_above_cutoff <- subset(TSS_positions_unfiltered, TSS_trans_ratio[TSS_positions_unfiltered]>TSS_trans_ratio_cutoff)
TSS_positions_sorted<- base::strsplit(TSS_above_cutoff, "-")
TSS_nonReduced_Ranges <- create_GRanges_object(TSS_positions_sorted, res_PSSTSS[TSS_above_cutoff,]$baseMean)
rm(trans_norm)
rm(TSS_norm)
```

## Create .gff files

```{r create-gff}
PSS_positions_Ranges <- GenomicRanges::reduce(PSS_nonReduced_Ranges)
length(PSS_positions_Ranges)

TSS_positions_Ranges <- GenomicRanges::reduce(TSS_nonReduced_Ranges)
length(TSS_positions_Ranges)

TSS_positions_Ranges$type <- "TSS"
PSS_positions_Ranges$type <- "PSS"
save_gff(c(TSS_positions_Ranges, PSS_positions_Ranges), "Output/gffs/", "TSS_PSS_multireads")
rm(TSS_positions_Ranges)
rm(PSS_positions_Ranges)
```

# Perform DESeq2 analysis for PSS

```{r PSS-perform-DESeq2}
PSS_use <- PSS_coverage[PSS_positions_unfiltered,]
rm(PSS_coverage)

coldata <- read.csv("Input/colData_PSS.csv", row.names=1)

ddsMat_PSS <- DESeqDataSetFromMatrix(countData = PSS_use,
                                 colData = coldata,
                                 design = ~ strain)

ddsMat_PSS$sizeFactor <- PSS_fact
ddsMat_PSS <- DESeq(ddsMat_PSS) 
rm(PSS_use)
```

## Diagnostic Plots

```{r}
plotDispEsts(ddsMat_PSS, main="PSS comparison", xlab="Mean of Normalized Counts", ylab="Dispersion")
p <- PCA_plot(ddsMat_PSS, "PSS")
p 
ggsave("Output/DESeq2_Plots/ddsMat_PSS_PCA.pdf", plot=p, width=9, height=9, units="cm")
p <- heatmap_plot(ddsMat_PSS, "PSS")
p
ggsave("Output/DESeq2_Plots/ddsMat_PSS_heatMap.pdf", plot=p, width=15, height=12, units="cm")
```

## Extract Results

```{r}
# extract results
PSS_result_dWT_WT <- results(ddsMat_PSS, contrast=c("strain", "dWT", "WT")) # dWT/WT -> higher in dWT: higher log2FC
write.csv(PSS_result_dWT_WT[order(PSS_result_dWT_WT$padj),], file="Output/DESeq2_resultsTables/results_PSS-dWT-WT.csv")

PSS_result_dWT_TV <- results(ddsMat_PSS, contrast=c("strain", "dWT", "TV")) # dWT/TV -> higher in dWT: higher log2FC
write.csv(PSS_result_dWT_TV[order(PSS_result_dWT_TV$padj),], file="Output/DESeq2_resultsTables/results_PSS-dWT-TV.csv")

PSS_result_WT_TV <- results(ddsMat_PSS, contrast=c("strain", "WT", "TV")) # WT/TV -> higher in WT: higher log2FC
write.csv(PSS_result_WT_TV[order(PSS_result_WT_TV$padj),], file="Output/DESeq2_resultsTables/results_PSS-WT-TV.csv")

rm(ddsMat_PSS)
```

```{r}
indices_plus <- which(strand(PSS_nonReduced_Ranges)=="+")
indices_minus <- which(strand(PSS_nonReduced_Ranges)=="-") 
PSS_names <- c()
PSS_names[indices_plus] <- paste(seqnames(PSS_nonReduced_Ranges[indices_plus]),start(PSS_nonReduced_Ranges[indices_plus]),"plus",sep="-")
PSS_names[indices_minus] <- paste(seqnames(PSS_nonReduced_Ranges[indices_minus]),start(PSS_nonReduced_Ranges[indices_minus]),"minus",sep="-")
rm(indices_plus)
rm(indices_minus)
PSS_nonReduced_Ranges$names <- PSS_names
PSS_nonReduced_Ranges$featuresOverlap <- rep("", length(PSS_nonReduced_Ranges))
PSS_nonReduced_Ranges$TU_overlap <- rep("", length(PSS_nonReduced_Ranges))
for(i in 1:length(PSS_nonReduced_Ranges)){
  PSS_nonReduced_Ranges$featuresOverlap[i] <- paste(features[which(countOverlaps(features,PSS_nonReduced_Ranges[i])>0)]$locus_tag,collapse=",")
  PSS_nonReduced_Ranges$TU_overlap[i] <- paste(TUs[which(countOverlaps(TUs,PSS_nonReduced_Ranges[i])>0)]$index,collapse=",")
}
```

```{r}
# create tables with annotation: which features and TUs are overlapping with PSS?
df_PSS_nonRed_Ranges <- as.data.frame(PSS_nonReduced_Ranges)
row.names(df_PSS_nonRed_Ranges) <- df_PSS_nonRed_Ranges$names

# PSS result dWT WT
PSS_result_dWT_WT_annot <- rownames_to_column(as.data.frame(PSS_result_dWT_WT[order(PSS_result_dWT_WT$padj),]))
PSS_result_dWT_WT_annot$CDS <- df_PSS_nonRed_Ranges[PSS_result_dWT_WT_annot$rowname,]$featuresOverlap
PSS_result_dWT_WT_annot$TUs <- df_PSS_nonRed_Ranges[PSS_result_dWT_WT_annot$rowname,]$TU_overlap
PSS_result_dWT_WT_annot <- PSS_result_dWT_WT_annot[,c(8,9,1,2:7)]
write_tsv(PSS_result_dWT_WT_annot, "Output/DESeq2_resultsTables/results_PSS_dWT_WT_annotated.tsv")

# PSS result dWT TV
PSS_result_dWT_TV_annot <- rownames_to_column(as.data.frame(PSS_result_dWT_TV[order(PSS_result_dWT_TV$padj),]))
PSS_result_dWT_TV_annot$CDS <- df_PSS_nonRed_Ranges[PSS_result_dWT_TV_annot$rowname,]$featuresOverlap
PSS_result_dWT_TV_annot$TUs <- df_PSS_nonRed_Ranges[PSS_result_dWT_TV_annot$rowname,]$TU_overlap
PSS_result_dWT_TV_annot <- PSS_result_dWT_TV_annot[,c(8,9,1,2:7)]
write_tsv(PSS_result_dWT_TV_annot, "Output/DESeq2_resultsTables/results_PSS_dWT-TV_annotated.tsv")

# PSS result WT, TV
PSS_result_WT_TV_annot <- rownames_to_column(as.data.frame(PSS_result_WT_TV[order(PSS_result_WT_TV$padj),]))
PSS_result_WT_TV_annot$CDS <- df_PSS_nonRed_Ranges[PSS_result_WT_TV_annot$rowname,]$featuresOverlap
PSS_result_WT_TV_annot$TUs <- df_PSS_nonRed_Ranges[PSS_result_WT_TV_annot$rowname,]$TU_overlap
PSS_result_WT_TV_annot <- PSS_result_WT_TV_annot[,c(8,9,1,2:7)]
write_tsv(PSS_result_WT_TV_annot, "Output/DESeq2_resultsTables/results_PSS_WT-TV_annotated.tsv")
```

Run in respective Directory (output/DESeq2_resultsTables/)
python    changeAnnotation_DESeq2.py results_PSS_dWT_WT_annotated.tsv results_dWT_WT_annotated-2.tsv
python    changeAnnotation_DESeq2.py results_PSS_dWT-TV_annotated.tsv results_dWT-TV_annotated-2.tsv
python    changeAnnotation_DESeq2.py results_PSS_WT-TV_annotated.tsv results_PSS-WT-TV_annotated-2.tsv

```{r, fig.wide = TRUE}
pdf(file="Output/DESeq2_Plots/ddsMat_PSS-dWT-WT_pValuePlot.pdf", width=4.5, height=4.5)
pvaluePlot(PSS_result_dWT_WT, "PSS dWT WT")
dev.off()
pdf(file="Output/DESeq2_Plots/ddsMat_PSS-dWT-TV_pValuePlot.pdf", width=4.5, height=4.5)
pvaluePlot(PSS_result_dWT_TV, "PSS dWT dTV")
dev.off()
pvaluePlot(PSS_result_dWT_WT, "PSS dWT WT")
pvaluePlot(PSS_result_dWT_TV, "PSS dWT dTV")
```

Filter out PSS below filter level of results(). For these, p.adj is set to NA - hence: remove positions at which p.adj==NA

```{r}
PSS_result_dWT_WT <- subset(PSS_result_dWT_WT, !is.na(PSS_result_dWT_WT$padj))
PSS_result_dWT_TV <- subset(PSS_result_dWT_TV, !is.na(PSS_result_dWT_TV$padj))
PSS_result_WT_TV <- subset(PSS_result_WT_TV, !is.na(PSS_result_WT_TV$padj))
```

### PSS dWT WT 

```{r, fig.wide = TRUE}
count_up_down(PSS_result_dWT_WT, foldchange=1, padjusted=0.05)

p <- volcanoPlot_ggplot(as.data.frame(PSS_result_dWT_WT), foldchange=1, padjusted=0.05) + scale_colour_manual(values=c("DOWN"="#000000ff", "UP"="#005a96ff", "NO"="#d3d3d3ff"))
p
ggsave("Output/DESeq2_Plots/ddsMat_PSS-0h_VolcanoPlot.pdf", plot=p, width=15, height=12, units="cm")

p <- MAplot_ggplot(PSS_result_dWT_WT, foldchange=1.0, y_axis_label = "Log2 fold-change(rne(WT)/WT)") + scale_colour_manual(values=c("DOWN"="#000000ff", "UP"="#005a96ff", "NO"="#d3d3d3ff"))
p
ggsave("Output/DESeq2_Plots/ddsMat_PSS-dWT-WT_MAplot.pdf",plot=p, width=15, height=12, units="cm")
```

### PSS dWT-TV

```{r, fig.wide = TRUE}
count_up_down(PSS_result_dWT_TV, foldchange=1, padjusted=0.05)

p <- volcanoPlot_ggplot(as.data.frame(PSS_result_dWT_TV), foldchange=1, padjusted=0.05) + scale_colour_manual(values=c("DOWN"="#e69f00ff", "UP"="#005a96ff", "NO"="#d3d3d3ff"))
p
ggsave("Output/DESeq2_Plots/ddsMat_PSS-dWT-TV_VolcanoPlot.pdf", plot=p, width=15, height=12, units="cm")

p <- MAplot_ggplot(PSS_result_dWT_TV, foldchange=1.0, y_axis_label = "Log2 fold-change(rne(WT)/rne(5p))") + scale_colour_manual(values=c("DOWN"="#e69f00ff", "UP"="#005a96ff", "NO"="#d3d3d3ff"))# + ylim(-5,+5)
p
ggsave("Output/DESeq2_Plots/ddsMat_PSS-1h_MAplot.pdf",plot=p, width=15, height=12, units="cm")
```

### PSS WT-TV

```{r}
count_up_down(PSS_result_WT_TV, foldchange = 1, padjusted = 0.05)
  
p <- MAplot_ggplot(PSS_result_WT_TV, foldchange=1.0, y_axis_label = "Log2 fold-change(WT/rne(5p))") + scale_colour_manual(values=c("DOWN"="#e69f00ff", "UP"="#000000ff", "NO"="#d3d3d3ff"))# + ylim(-5,+5)
p
ggsave("Output/DESeq2_Plots/ddsMat_PSS-WT-TV_MAplot.pdf",plot=p, width=15, height=12, units="cm")
```

```{r, echo=FALSE}
save.image(file = "multireads.RData")
```
```{r, echo=FALSE}
#load("multireads.RData")
```

# Session info {.unnumbered}

```{r sessionInfo, echo=FALSE}
sessionInfo()
```
